---
title: "SFC GM Workshop"
author: "Corinthia Black"
date: "10/26/2020"
output: html_document
---

## Libraries
   This block contains the libraries you need to preform the following:
      landmarking = StereoMorph
                    (https://aaronolsen.github.io/software/stereomorph.html)
      analysis = geomorph
                    (https://cran.r-project.org/web/packages/geomorph/geomorph.pdf)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("StereoMorph") #landmarking program
library(StereoMorph)
#install.packages("geomorph")
library(geomorph)
```

## Landmark Specimens
```{r Landmarking}
## Digitize Images
  # Click box 'Automatically save when advancing to next image' under Settings tab
  # Use X to place landmark, N to advance to next landmark, D to delete selected landmark
    # digitizeImages() with the following parameters: image.file (an image or folder of images to be digitized) and shapes.file (a file or folder where the digitized data will be saved)

digitizeImages(image.file ="Example_Dataset", shapes.file = "Landmarked", landmarks.ref="landmarks.txt")
```

## Analysis
You can easily read the shapes files into R using the readShapes() function in StereoMorph
```{r Read shapes}
# Read all shape data from all files
  # WINDOWS USERS use shapes=readShapes('Landmarked')
shapes=readShapes('Landmarked/')

# readShape() produces a list with several elements
  # Landmarks are saved as landmarks.pixel OR landmarks.scaled (either matrix will work)
  # We will save the landmarks as it's own element for futher analyses
landmarks=shapes$landmarks.scaled

# This step checks to see if there are any missing landmark data. If so, you can go back and fix the data, or use estimate.missing() in geomorph if needed
any(is.na(landmarks))
```

This chunk will run the GPA and estimate the meanshape of all our data (to be used later)
```{r GPA} 
# use gpagen() in geomorph to correct for orientation, translation and size
gpa=gpagen(landmarks) #saves as large list (coords = shape files; Csize = centroid size)

# plot your superimposed specimens
  # this step can be used to find outliers which may need landmarks corrected
plot(gpa)

# estimate the mean shape of all data
  # need mean data to visualize shape change across components
meanshape=mshape(gpa$coords)
plot(meanshape) #NOTE: this function returns a mean shape with numbers in place of landmark names. These numbers will be used to create a links file that can outline the shape

# Use landmark numbers to generate a file that will link landmarks together with a line (this file is already provided for you)
links=read.csv("links.csv", header = F)
plotAllSpecimens(gpa$coords, links=links)
```

```{r PCA}
# We will use classifier information to produce colored PCAs and in statistics
  # in this case we are comparing stream and reservoir fishes
classifier=read.csv("classifiers.csv", header=T, row.names = 1)

# Preform the PCA using the GPA landmark coordinates
PCA=gm.prcomp(gpa$coords)

# To find out the importance of each component
summary(PCA)

# Generic plot of first 2 components
plot(PCA)

# Let's add colors for the different groups and labels
plot(PCA, pch=22, cex=1.5, bg=as.numeric(classifier$Location))
text(PCA$x, labels = labels(PCA$x)[[1]], pos=3) #change position of text with pos (1=under, 2=left, 3=above, 4=right)

# To view what the shape changes are along the component use plotRefToTarget
  # The first shape listed will be grey and the last will be black
  # Shapes are found in the PCA list (see example below)
plotRefToTarget(meanshape, PCA$shapes$shapes.comp1$min, method = "points", links = links) # mean shape vs negative end of PC1
#method = "TPS", "vector", "points", or "surface" (surface only works with 3D data) try the different methods

plotRefToTarget(meanshape, PCA$shapes$shapes.comp1$max, method = "points", links = links) # mean shape vs positive end of PC1

plotRefToTarget(PCA$shapes$shapes.comp1$min, PCA$shapes$shapes.comp1$max, method = "points", links = links) #negative end of PC1 vs postitive end of PC1
```

PCA is a great way to visualize differences, but we only see part of the story. If you look at the other components, there is a ton of shape data that is unaccounted for in the PCA plots.
We could plot all the PCs in 2- or 3-dimensions, but we can also use statistics to see if the total shape is actually different.

To do that, we will use a procrusties ANOVA
```{r ANOVA}
# To run a procrusties ANOVA we need a linear model. Our hypothesis wants to know if there are shape differences between stream fish and reservoir fish. We are also interested in if size may be a factor.
  # can change sum of squares type: type I (sequential), type II (hierarchical), or type III (marginal) sums of squares and cross-products computation - good summary here http://www.utstat.utoronto.ca/reid/sta442f/2009/typeSS.pdf

fit.location=procD.lm(gpa$coords ~ classifier$Location, SS.type = "III")
summary(fit.location)

fit.complex=procD.lm(gpa$coords ~ classifier$Location + gpa$Csize + classifier$Location*gpa$Csize, SS.type = "III")
summary(fit.complex)

# F-drop Test: is p < 0.05 = complex model is best
  # more complex model goes 1st followed by less complex
anova(fit.location, fit.complex)
```

Size might also account for shape variation (small fish are slim vs large fish are more rotund)

To analyze this, we can use allometric analyses
```{r Allometric analysis}
# To test for allometry, we use the same format as the procrusties ANOVA!
fit.size=procD.lm(gpa$coords ~ gpa$Csize, SS.type = "III")
summary(fit.size)

# Plot prediction and Regression score
  # RegScore calculates a regression score as a projection of data on normalized vector that expresses the covariation between shape and the regression coefficients for size, conditioned on other model effects
    # For a single allometry, these shape scores are mathematically identical to the CAC [common allometric component of the shape data] (Adams et al. 2013).
  # Prediction line (PredLine) plots the first PC of fitted values against size
#http://search.r-project.org/library/geomorph/html/plotAllometry.html
#https://rdrr.io/cran/geomorph/man/shape.predictor.html

#Regression score
reg.plot = plot(fit.size, type="regression", reg.type = "RegScore", predictor = log(gpa$Csize), bg=as.numeric(classifier$Location), pch=22, cex=1.5) # make sure to have a predictor 

preds = shape.predictor(fit.size$GM$fitted, x= reg.plot$RegScore, Intercept=F, 
                         predmin = min(reg.plot$RegScore),
                         predmax = max(reg.plot$RegScore))

plotRefToTarget(meanshape, preds$predmin, mag=3, method = "points", links = links)
plotRefToTarget(meanshape, preds$predmax, mag=3, method = "points", links = links)


# Prediction line
pred.plot = plot(fit.size, type="regression", reg.type = "PredLine", predictor = log(gpa$Csize), bg=as.numeric(classifier$Location), pch=22, cex=1.5)
```